\chapter{Exception Handling}\label{chap:exception}
\index{exceptions|(}

We define the term {\em exceptions} as errors in program execution
that are handled by a non-local change in execution state.  Exception
handling in XSB is ISO-compatible, and has been extended to handle
tabled evaluations.  

\section{The Mechanics of Exception Handling}
%
We address the case of non-tabled evaluations before discussing the
extensions for tabling.

\subsection{Exception Handling in Non-Tabled Evaluations}
%
The preferred mechanism for dealing with exceptions in XSB is to use
the predicates {\tt catch/3} and {\tt default\_user\_error\_handler/1}
together with one of XSB's error predicates (such as {\tt
  misc\_error/1}).  These predicates are ISO-compatible, and their use
can give a great deal of control to exception handling.  At a high
level, when an exception is encountered an error term $T$ is {\em
  thrown}.  In a non-tabled Prolog program, throwing an error term $T$
causes XSB to examine its choice point stack until it finds a {\em
  catcher} that unifies with $T$.  This catcher then calls a {\em
  handler}.  If no explicit catcher for $T$ exists, a default handler
is invoked, which usually results in an abort, and returns execution
to the top-level of the interpreter, or to the calling C
function.\footnote{Starting in Version 3.5.1, XSB uses the ISO
  compliant {\tt error/2} for error terms, rather than {\tt error/3}
  as in previous versions.}

A handler is set up when {\tt catch(Goal,Catcher,Handler)} is called.
At the time of the call, a continuation is saved (i.e. a Prolog choice
point), and {\tt Goal} is called.  If no exceptions are encountered,
answers for {\tt Goal} are obtained as usual.  However, within the
execution of {\tt Goal}, an exception might be thrown by calling a
Prolog predicate in the {\tt error\_handler} module, or by executing a
C-level error function.~\footnote{A user-defined error type is
  desired, the Prolog predicate {\tt throw/1} can also be called
  directly.}  As mentioned above, when an error is thrown in an
environment $Env$, XSB searches for an ancestor $Env_{anc}$ of $Env$
in which {\tt catch/3} was called, and in which the catcher (second
argument) unifies with {\tt Error}.  If such an ancestor is found,
program execution reverts to the ancestor and all intervening choice
points are removed.  The catcher's {\tt Handler} goal is called and
the exception is thereby handled.  On the other hand, if no ancestor
in the user's program was called using {\tt catch/3} the exception is
handled via the handler associated with XSB's goal interpreter at the
top-level command line or C API.  This top-level handler checks
whether a clause with head {\tt default\_user\_error\_handler(Term)}
has been asserted, such that {\tt Term} unifies with {\tt Error}.  If
so, this handler is executed.  If not, XSB's default system error
handler in invoked an error message is output and execution returns to
the top level of the interpreter.

The following, somewhat fanciful, example helps clarify these
concepts~\footnote{Code for this example can be found in {\tt
\$XSBDIR/examples/exceptions.P}.}.  Consider the predicate {\tt
userdiv/2} (Figure~\ref{fig:userdiv}) which is designed to be called
with the first argument instantiated to a number.  A second number is
then read from a console, and the first number is divided by the
second, and unified with the second argument of {\tt userdiv/2}.  By
using {\tt catch/3} and {\tt throw/1} together the various types of
errors can be caught.

%------------------------------------------------------------------------------------------
\begin{figure}[hbtp]
\longline
\begin{small}
\begin{verbatim}
:- import error_writeln/1 from standard.
:- import type_error/4 from error_handler.

userdiv(X,Ans):- 
        catch(userdiv1(X,Ans),mydiv1(Y),handleUserdiv(Y,X)).

userdiv1(X,Ans):- 
        (number(X) -> true; type_error(number,X,userdiv1/2,1)),
        write('Enter a number: '),read(Y),
        (number(Y) -> true ; throw(mydiv1(error1(Y)))),
        (Y < 0 -> throw(mydiv1(error2(Y))); true),
        (Y =:= 0 -> throw(error(zerodivision,userdiv/1,[])); true),
        Ans is X/Y.

handleUserdiv(error1(Y),_X):- 
        error_writeln(['a non-numeric denominator was entered in userdiv/1: ',Y]),fail.
handleUserdiv(error2(Y),_X):- 
        error_writeln(['a negative denominator was entered in userdiv/1: ',Y]),fail.
\end{verbatim}
\end{small}
\caption{The {\tt userdiv/1} program} \label{fig:userdiv}
\longline
\end{figure}
%------------------------------------------------------------------------------------------

The behavior of this program on some representative inputs is shown
below.

\begin{small}
\begin{verbatim}
| ?- userdiv(p(1),F).
++Error[XSB/Runtime/P]: [Type (p(1) in place of number)]  in arg 1 of predicate userdiv1/2
Forward Continuation...
... machine:xsb_backtrace/1
... error_handler:type_error/4
... standard:call/1
... x_interp:_$call/1
... x_interp:call_query/1
... standard:call/1
... standard:catch/3
... x_interp:interpreter/0
... loader:ll_code_call/3
... standard:call/1
... standard:catch/3

no
| ?- userdiv(3,F).
Enter a number: foo.
a non-numeric denominator was entered in userdiv/1: foo

no
|| ?- userdiv(3,F).
Enter a number: -1.
a negative denominator was entered in userdiv/1: -1

no
| ?- userdiv(3,Y).
Enter a number: 2.

Y = 1.5000

yes
\end{verbatim}
\end{small}

\noindent
Note, however the following behavior.

\begin{small}
\begin{verbatim}
| ?- userdiv(3,F).
Enter a number: 0.
++Error[XSB/Runtime/P] uncaught exception: error(zerodivision,userdiv / 1)
Aborting...
\end{verbatim}
\end{small}

\noindent
By examining the program above, it can be seen that if {\tt p(1)} is
entered, the predicate {\tt type\_error/3} is called.  {\tt
  type\_error/3} is an XSB mechanism to throw a type error from
Prolog.  Error terms thrown by system predicates such as {\tt
  type\_error/3} are in XSB's {\em standard error format}, which is
ISO-compatable and which may encode useful information.  For instance,
the type error thrown in the above example is known to XSB's default
system error handler which prints out a message along with a {\em
  backtrace} that indicates the calling context in which the error
arose (this behavior can be controlled: see
Section~\ref{sec:backtrace}).  Alternately, in the second case, when
{\tt -1} is entered, the (non-standard) error term {\tt
  mydiv1(error2(-1))} is thrown, which is caught within {\tt
  userdiv/2} and handled by {\tt handleUserdiv/2}.  Finally, when {\tt
  0} is entered for the denominator, an error term of the form {\tt
  error(zerodivision,userdiv/1)} is thrown, and this term does not
unify with the second argument of the {\tt catch/3} literal in the
body of {\tt userdiv/1}, or with any error in standard format.  The
error is instead caught by XSB's default system error handler which
prints an uncaught exception message and aborts to the top level of
the interpreter.

XSB has two default system error handlers: one used when XSB is called
as a stand-alone process, and another when XSB is embedded in a
process.  Each recognizes the same error formats (see Section
\ref{sec:iso-errors}), and handles the rest as uncaught exceptions.
However, there may be times when an application requires special
default handling: perhaps the application calls XSB from through a
socket, so that aborts are not practical.  As another example, perhaps
XSB is being called from a graphical user interface via
InterProlog~\cite{Cale01} or some other interface, so that in addition
to a special abort handling, one would like to display an error
window.  In these cases it is convenient to make use of the dynamic
predicate {\tt default\_user\_error\_handler/1}.  {\tt
  default\_user\_error\_handler/1} is called immediately before the
default system error handler, and after it is ascertained that no
catcher for an error term is available via a {\tt catch/3} ancestor.

It is important to note that the system error handlers catch errors
only in the main thread, and do not affect errors thrown by goals
executed by {\tt thread\_create/[2,3]}.  Error terms thrown by goals
executed by non-detached threads are stored internally, and can be
obtained by {\tt thread\_join/2}.
%Error terms thrown by detached
%threads are lost when the thread exits, so that any error handling for
%a detached thread should be performed within the thread itself.  See
%Chapter~\ref{chap:threads} for further information.

Accordingly, suppose the following clause is asserted into {\tt
usermod}:
%
\begin{small}
\begin{verbatim}
?- assert((default_user_error_handler(error(zerodivision,Pred)):- 
        error_writeln(['Aborting: division by 0 in: ',Pred]))).
\end{verbatim}
\end{small}
%
The behavior will now be
\begin{small}
\begin{verbatim}
| ?- userdiv(4,F).
Enter a number: 0.
Aborting: division by 0 in: userdiv / 1
\end{verbatim}
\end{small}
The actions of {\tt catch/3} and {\tt throw/1} resemble that of the
Prolog cut in that they remove choice points that lie between a call
to {\tt throw/1} and the matching {\tt catch/3} that serves as its
ancestor. 

% However, if this process encounters a choice point for an
%incomplete table, execution is aborted to the top user level.

The predicate {\tt call\_cleanup/2}
(cf. Section~\ref{meta_predicates}) can be used with {\tt catch/3},
since the goal {\tt call\_cleanup(Goal,Cleanup)} executes {\tt
  Cleanup} whenever computation of {\tt Goal} is completed, whether
because {\tt Goal} has thrown an exception, has failed, or has
succeeded with its last answer.  {\tt call\_cleanup/2} can thus be
used to release resources created by {\tt Goal} (such as streams,
mutexes, database cursors, etc.).  However, if {\tt Goal} throws an
exception, {\tt call\_cleanup/2} will re-throw the exception after
executing cleanup.

\index{tabling!and exceptions}
\index{strongly connected components (SCCs)}
\subsection{Exception Handling in Tabled Evaluation} \label{sec:exceptions-w-tables}
%
The exception handling as previously described requires extensions in
order to work well with tabled predicates.  First, if an {\em
  unhandled} exception is thrown duing evaluation of a tabled subgoal
$S$ and $S$ is not completed, the table for $S$ is not meaningful and
should be removed.  (Tables that have been completed are not affected
by exceptions.)  Accordingly, the user will sometimes see the message:
%
\begin{verbatim}
   Removing incomplete tables...
\end{verbatim}
%
written to standard feedback.  But what about exceptions that are {\em
  caught} during the computation of $S$?

The proper action to take in such a case is complicated by the
scheduling mechanism of tabling which, as discussed in
Chapter~\ref{chap:TablingOverview}, is more complex than in Prolog.
Rather than a simple depth-first search, as in Prolog, tabled
evaluations effectively perform a series of fixed-point computations
for various sets of mutually dependent subgoals, which are termed {\em
  SCCs}~\footnote{This term is used since sets of mutually dependent
  subgoals are formally modelled as (approximate) {\em Strongly
    Connected Components} within a dependency graph.}.  In fact, a
tabled evaluation can be seen as a tree of SCCs (in batched
evaluation) or a chain of SCCs (in local evaluation).  In a tabled
evalution XSB's throw mechanism searches for the nearest catcher $C$
among its ancestors
%
\begin{itemize}
\item whose first argument unifies with the thrown error; and
\item where $C$ is between SCCs: that is where the set of subgoals
  that depend on $C$ is disjoint from the set of subgoals upon which
  $C$ depends.  We term this the {\em SCC restriction} for exception
  handling.
\end{itemize}
%

This behavior can be best understood by an example.  Consider the
query {\tt a(X)} to the program in Figure~\ref{fig:tab_except} which
has the following output:
%
%-------------------------------------------------------------------------------
\begin{figure}[bhtp]
\longline
\begin{small}
\begin{verbatim}
:- table a/1, b/1, c/1,d/1.
a(X):- writeln(a_calling_b),b(X).

b(X):- writeln(b_calling_a),a(X).
b(X):- writeln(b_calling_c),catch(c(X),_,(writeln(handled_1),fail)).

c(X):- writeln(c_calling_d),d(X).
c(X):- writeln(c_aborting),abort.
d(X):- writeln(d_calling_c),catch(c(X),_,(writeln(handled_2),fail)).
\end{verbatim}
\end{small}
\caption{A program to illustrate exception handling in tabled evaluations} \label{fig:tab_except}
\longline
\end{figure}
%------------------------------------------------------------------------------

\begin{small}
\begin{verbatim}
   | ?- a(X).
   a_calling_b
   b_calling_a
   b_calling_c
   c_calling_d
   d_calling_c
   c_aborting
   Removing incomplete tables...
   handled_1
\end{verbatim}
\end{small}
%
Note that there are 2 SCCs, $\{a(X),b(X)\}$ and $\{c(X),d(X)\}$.  When
the {\tt abort} is called in the body of {\tt c(X)} the catch in the
body of {\tt d(X)} is its nearest ancestor; however this catch is
skipped over, and the catch in the body of {\tt b(X)} takes effect.
This catch is between the SCCs -- the first SCC depends on it, but the
second doesn't.
%
Due to the SCC restriction, the actual behavior of exception handling
with tabling is thus somewhat less intuitive than in Prolog.  If this
restriction were lifted, there would be no guarantee that there
existed a unique catch that was the closest ancestor of an exception.

While the above mechanism offers a great deal of flexibility, for many
cases the best approach to exception handling is to keep it simple.
%
\begin{enumerate}
\item Use catches when there will be no tabled subgoal between an
  exception and its catcher.  For instance, sometimes it may be
  annoying to have {\tt atom\_codes/2} throw an exception rather than
  failing, if given an integer in its first argument.  This can be
  addressed by the predicate
%
\begin{verbatim}
my_atom_codes(X,Y):- 
   catch(atom_codes(1,B),error(type_error(A,B),C,D),writeln(E)).
\end{verbatim}
%
which, for a type error, does not interact with tabling in any way.  
%
\item Similarly, if only subgoals to {\em completed} tables occur
  between an exception and its catcher, exception handling behaves
  just as in case 1).
%
\item Otherwise, abort the entire tabled computation and handle it
  from there.  (Unless you really know what you're doing!)
\end{enumerate}
%

\index{Prolog flags!{\tt exception\_pre\_action}}
\index{strongly connected components (SCCs)}
\predref{print\_sdg\_subgoal\_info}
\subsubsection{Obtaining Information about a Tabled Computation after an Exception is Thrown}
%
XSB backtraces (Section~\ref{sec:backtrace}) provide information about
the context in which error is thrown, but in a tabled computation
additional information is available.  If the Prolog flag {\tt
  exception\_pre\_action} is set to {\tt print\_incomplete\_subgoals}
(its default setting is {\tt none}), then when an exception is thrown,
incomplete tables and their SCC information at the time an exception
is thrown are printed to a file.  The file may be obtained through the
predicate {\tt get\_scc\_dumpfile/1} in the module {\tt tables}.  No
file is generated unless the exception is thrown over at least one
incomplete table.

\section{XSB's Standard Format for Errors} \label{sec:iso-errors}
\index{ISO!errors}

All exceptions that occur during the execution of an XSB program can
be caught.  However, by structuring error terms in a consistent
manner, different classes of errors can be handled much more easily by
handlers, both system- and user-defined.  This philosophy partly
underlies the ISO Standard for defining classes of Prolog errors
\cite{ISO-Prolog}.  While the ISO standard defines various types of
errors and how they should arise during execution of ISO Prolog
predicates, it only partially defines the actual error terms a system
should use.  The ISO format can be represented as:

\begin{center}
{\tt error(Tag,Context), }
\end{center}

\noindent
where {\tt Tag} is specific to each class of error, while {\tt
  Context} is implementation-dependent.\footnote{If a program catches
  errors itself, {\tt error/2} may need to be imported from {\tt
    error\_handler}.}

\subsection{Error Tags}

In XSB, the ISO-compliant values for {\tt Tag} are given below.

\begin{description}
\item[{\tt domain\_error(Valid\_type,Culprit)}] is the tag for an ISO
  domain error, where {\tt Valid\_type} is the domain expected and
  {\tt Culprit} is the term observed.  Various ISO predicates may have
  specific domains for input values; and in addition unlike types,
  domains can be user-defined.
%
\item[{\tt evaluation\_error(Flag)}] is the tag for an ISO evaluation
  error (e.g. overflow or underflow), and {\tt Flag} is the type of
  evaluation error encountered (e.g., {\tt undefined}, if an
  arithmetic function is undefined for a given input).
%
\item[{\tt existence\_error(Type,Culprit)}] is the tag for an ISO
  existence error, where {\tt Type} is the type of a resource (e.g., a
  predicate, stream, attribute handler, etc.) and {\tt Culprit} is the
  term observed.
%
\item[{\tt instantiation\_error}] is the tag for an ISO instantiation
  error.
%
\item[{\tt permission\_error(Op,Obj\_type,Culprit)}] is the tag for an ISO
  permission error, when an operation {\tt Op} was applied to an
  object of type {\tt Obj\_type}, but {\tt Culprit} was observed.
%
\item[{\tt representation\_error(Flag)}] is the tag for an ISO
  representation error (e.g., the maximum arity of a predicate has
  been exceeded), and {\tt Flag} is the type of representation error
  encountered.
%
\item[{\tt resource\_error(Flag)}] is the tag for an ISO resource error
  (e.g. allowed memory has been used, or too many files have been
  opened), and {\tt Flag} is the type of resource error encountered.
%
\item {\tt syntax\_error} and {\tt syntax\_error(Culprit)} are alternate
  tags for an ISO syntax error, where {\tt Culprit} denotes a
  syntactically-incorrect sequence of tokens.
%
\item {\tt system\_error(Flag)} is the tag for an ISO system
  error, and {\tt Flag} is the type of system error encountered.
%
\item {\tt type\_error(Valid\_type,Culprit)} is the tag for an
  ISO type error, where {\tt Valid\_type} is the type expected and
  {\tt Culprit} is the term observed.  As opposed to domain errors,
  type errors should be used for checks of Prolog types only
  (i.e. integers, floats, atoms, etc.)
%
\end{description}

\noindent
In addition, XSB also makes use of some other classes of errors.
%
\begin{description}
\item[{\tt table\_error}] and {\tt type\_error(Subtype)} are the tags for an
  error arising when using XSB's tabling mechanism, when the condition
  giving rise to the error does npt easily fit under one of the above
  classes.
%
%\item[{\tt error(thread\_cancel,Id)}] is the format of an error ball for a
%  thread that has been cancelled by XSB thread {\tt Id} (See
%  Chapter~\ref{chap:threads} for details on thread cancellation.)
%
\item[{\tt misc\_error}] is the tag for an error that is not
  otherwise classified.
%
\item[{\tt misc\_error(Level)}] is the tag for {\tt abort/0-1}, and is
  used to allow users to abort a suspended query.  {\tt Level} is the
  break level at which the error was thrown.
%
\end{description}

In \version{} of XSB, errors for ISO predicates are usually, but
always ISO-compliant (we still have a few non-compliant errors to
catch).  However, when XSB determines it is out of available system
memory, recovering from such an error may be difficult at best.
Accordingly the computation is aborted in the sequential engine, or
XSB exits in the multi-threaded engine~\footnote{This does not include
  overflowaing a memory limit specified by the flag {\tt
    max\_memory}.}.

\subsection{XSB-Specific Information in Error Terms}

XSB also encodes other information in error terms, which may vary with
the error thrown, the form in which XSB was compiled, and the version
of XSB.  In addition, the specifics of how the information is
represented may vary, so that this information should always be
represented through the access methods described in this section along
with Section~\ref{sec:catch}.

\begin{description}

\item[{\tt Message}] describes the error in human-readable format.
  Messages are present in all of XSB's system errors, and can be
  obtained through {\tt xsb\_error\_get\_message/2}.

\item[{\tt Goal}] represents tabled goal that is closest to the
  environment of the thrown error.  It is present in some, but not all
  error terms thrown by XSB and can be obtained as a term through {\tt
    xsb\_error\_get\_goal/2}, and as an atom through {\tt
    xsb\_error\_get\_goalatom/2}.

\item[{\tt Thread Id}] is an atom {\tt 'th <tid>'} indicating the id
  of the thread that threw the error.  Thread Id information is only
  present when using the multi-threaded version of XSB, and even in
  that version is not present in all error terms.  The predicate {\tt
    xsb\_error\_get\_tid/2} can be used to obtain this information if
  present.

\item[{\tt Backtrace}] represents the stack of the forward
  continuations in the execution stack at the time the error was
  thrown.  Backtraces are present by default in all XSB system error
  terms, and are described in Section~\ref{sec:backtrace}.  They may be
  obtained from an error term using the predicate {\tt
    xsb\_error\_get\_backtrace/2}.

\end{description}

%---------------------------------------------------------------------------------------
\section{Predicates to Throw and Handle Errors}
\label{sec:errorpredicates}

\subsection{Predicates to Throw Errors}

XSB provides a variety of predicates that throw errors~\footnote{C
  functions for throwing terms and ISO-style errors are described in
  Volume 2, Chapter 3 {\em Foreign Language Interface}.}. In general,
we recommend the use of predicates such as {\tt domain\_error/4} over
the direct use of {\tt throw/1} when possible.
%
\begin{description}
\isoitem{throw(+ErrorTerm)}{throw/1}
%
Throws the error {\tt ErrorTerm}.  Execution traverses up the choice
point stack until a goal of the form {\tt catch(Goal,Term,Handler)} is
found such that {\tt Term} unifies with {\tt ErrorTerm}.  In this
case, {\tt Handler} is called.  If no catcher is found in the main
thread, the system looks for a clause of {\tt
  default\_user\_error\_handler(Term)} such that {\tt Term} unifies
with {\tt ErrorTerm} --- if no such clause is found the default system
error handler is called.  In a non-main joinable thread, the error
term is stored internally and the thread exits; in a detached thread,
the thread exits with no action taken.  {\tt throw/1} is most useful
in conjunction with specialized handlers for new types of errors not
already supported in XSB.  
%
\ourmoditem{domain\_error(+Valid\_type,-Culprit,+Predicate,+Arg)}{domain\_error/4}{error\_handler}
%
Throws a domain error.  Using the default system error handler (with
the Prolog flag {\tt backtrace\_on\_error} set to off) an example is
{\small
\begin{verbatim}
domain_error(posInt,-1,checkPosInt/3,3).
++Error[XSB/Runtime/P]: [Domain (-1 not in domain posInt)] in arg 3 of predicate 
checkPosInt/3
\end{verbatim} }
%
\ourmoditem{evaluation\_error(+Flag,+Predicate,+Arg)}{evaluation\_error/3}{error\_handler}
%
Throws an evaluation error.  Using the default system error handler
(with the Prolog flag {\tt backtrace\_on\_error} set to off) an
example is {\small
\begin{verbatim}
evaluation_error(zero_divisor,unidiv/1,2).
++Error[XSB/Runtime/P]: [Evaluation (zero_divisor)] in arg 2 of predicate unidiv/2
\end{verbatim} }
%
\ourmoditem{existence\_error(+Object\_type,?Culprit,+Predicate,+Arg)}{existence\_error/4}{error\_handler}
%
Throws an existence error.  Using the default system error handler
(with the Prolog flag {\tt backtrace\_on\_error} set to off) an
example is {\small
\begin{verbatim}
existence_error(file,'myfile.P','load_intensional_rules/2',2).
++Error[XSB/Runtime/P]: [Existence (No file myfile.P exists)]  in arg 2 of predicate 
load_intensional_rules/2
\end{verbatim}
}
%
\ourmoditem{instantiation\_error(+Predicate,+Arg,+State)}{instantiation\_error/4}{error\_handler}
%
Throws an instantiation error.  Using the default system error
handler, an example (with the Prolog flag {\tt backtrace\_on\_error}
set to off) is {\small
\begin{verbatim}
?- instantiation_error(foo/1,1,nonvar).
++Error[XSB/Runtime/P]: [Instantiation]  in arg 1 of predicate foo/1: must be nonvar
\end{verbatim}
}
%
\ourmoditem{permission\_error(+Op,+Obj\_type,?Culprit,+Predicate)}{permission\_error/4}{error\_handler}
%
Throws a permission error.  Using the default system error handler, an
example (with the Prolog flag {\tt backtrace\_on\_error} set to off)
is {\small
\begin{verbatim}
| ?- permission_error(write,file,'myfile.P',foo/1).
++Error[XSB/Runtime/P]: [Permission (Operation) write on file: myfile.P]  in foo/1
\end{verbatim}
}
%
\ourmoditem{representation\_error(+Flag,+Predicate,+Arg)}{representation\_error/3}{error\_handler}
% 
Throws a representation error.  Using the default system error handler, an
example (with the Prolog flag {\tt backtrace\_on\_error} set to off) is {\small
\begin{verbatim}
representation_error(max_arity,assert/1,1).
++Error[XSB/Runtime/P]: [Representation (max_arity)] in arg 1 of predicate assert/1
\end{verbatim} }
%
\ourmoditem{resource\_error(+Flag,+Predicate)}{resource\_error/3}{error\_handler}
%
Throws a resource error.  Using the default system error handler
(with the Prolog flag {\tt backtrace\_on\_error} set to off) and example is {\small
\begin{verbatim}
resource_error(open_files,open/3)
++Error[XSB/Runtime/P]: [Resource (open_files)] in predicate open/3
\end{verbatim} }
%
\ourmoditem{type\_error(+Valid\_type,-Culprit,+Predicate,+Arg)}{type\_error/4}{error\_handler}
%
Throws a type error.  Using the default system error handler, an
example (with the Prolog flag {\tt backtrace\_on\_error} set to off)
is {\small
\begin{verbatim}
| ?- type_error(atom,f(1),foo/1,1).
++Error[XSB/Runtime/P]: [Type (f(1) in place of atom)]  in arg 1 of predicate foo/1
\end{verbatim}
}
%
\index{break level}
\ourmoditem{misc\_error(+Message)}{misc\_error/1}{error\_handler}
%
Throws a miscellaneous error that will be caught by the default system
handler at the current break level.  Usually, miscellaneous errors
should only be thrown when the cases above are not applicable, and the
type of error is not of interest for structured error handling.  Such
situations occur can occur for instance in debugging, during program
development. or for other reasons.  Conceptually, {\tt misc\_error/1}
differs from {\tt abort/0-1} only when called from a break level.
%
\repeatstandarditem{abort(+Message)}{abort/1}
\standarditem{abort}{abort/0}
%
Throws a type of miscellaneous error that will be caught by the
default system handler at the top level of the command-line intepreter
(CLI), but not at the current break level (if any).  This type of
exception can be useful, for instance if a long-running query is
interrupted by a ctrl-C, which suspends the query and starts a new
break level for XSB.  If a regular error is thrown, it will be caught
by the CLI for the current break level, and so does not provide a way
to abort the top-level query.  On the other hand, the exception thrown
by {\tt abort/0-1} will only be caught by the top-level CLI.  
%
\end{description}
%
% \ouritem{abort}\index{\texttt{abort/0}}
% Abandons the current execution and returns to the top level.  This
%     predicate should normally normally be used: 
% \begin{itemize} 
%\item when a non-ISO exception has occurred and the user wishes to
%abort the computation to the top-level of the interpreter.  
%
%\item {\em and} the type of the error is not of interest for
%structuring error handling.
%\end{itemize}
%
%Such situations occur can occur for instance in debugging, during
%program development, or in small-special purpose programs.
%
% \ouritem{abort(+Message)}\index{\texttt{abort/1}} \index{\texttt{STDERR}}
%    Acts as {\tt abort/0} but sents {\tt Message} to {\tt STDERR}
%    before aborting.

\subsection{Predicates used in Handling Errors}~\label{sec:catch}
%
For best results, output for handling errors should be sent to XSB's
standard error stream using the alias {\tt user\_error} or one of the
predicates described below.

\begin{description}
\isoitem{catch(?Goal,?CatchTerm,+Handler)}{catch/3}
%
Calls {\tt Goal}, and sets up information so that future throws will
be able to access {\tt CatchTerm} under the mechanism mentioned
above. {\tt catch/3} does not attempt to clean up system level
resources with the exception of incomplete tables, which are abolished
as discussed in Section~\ref{sec:exceptions-w-tables}.  However, it is
left up to the handler to close any open files, reset current input
and output, and so on \footnote{cf. the default system error handler,
  which performs these functions, if needed.}.
%
\standarditem{default\_user\_error\_handler(?CatchTerm)}{default\_user\_error\_handler/1}
%
Handles any error terms that unify with {\tt CatchTerm} that are not
caught by invocations of {\tt catch/3}.  This predicate closes open
tables and release mutexes held by the calling thread, but does not
attempt to clean up other system level resources, which is left to the
handler.
%

\index{streams!STDERR}
\ourrepeatmoditem{error\_write(?Message)}{error\_write/1}{standard}
\ourmoditem{error\_writeln(?Message)}{error\_writeln/1}{standard}
%
Utility routines for user-defined error catching.  These predicates
output {\tt Message} to XSB's {\tt STDERR} stream, rather than to
XSB's {\tt STDOUT} stream, as does {\tt write/1} and {\tt writeln/1}.
In addition, if {\tt Message} is a comma list, the elements in the
comma list are output as if they were concatenated together.  Each of
these predicates must be implicitly from the module {\tt standard}.

\ourmoditem{xsb\_error\_get\_message(Error,Message)}{xsb\_error\_get\_message/2}{error\_handler} 
%
Obtains the message associated with an error in XSB's standard format.
All errors in standard format have messages.

\ourmoditem{xsb\_error\_get\_goal(Error,?Goal}{xsb\_error\_get\_goal/2}{error\_handler}
%
Obtains the goal (represented as a Prolog term), if any, from an error term
that is in XSB's standard format.  If the error term has no goal, the
predicate fails.

\ourmoditem{xsb\_error\_get\_goalatom(Error,?GoalAtom}{xsb\_error\_get\_goalatom/2}{error\_handler}
%
Obtains the goal (represented as a Prolog atom), if any, from an error
term that is in XSB's standard format.  If the error term has no goal,
the predicate fails.  This routine is slightly more efficient than
{\tt xsb\_error\_get\_goal/2}.

\ourmoditem{xsb\_error\_get\_tid(Error,?Tid)}{xsb\_error\_get\_tid/2}{error\_handler} 
%
Obtains the atom {\tt 'th <tid>'} indicating the id of the thread that
threw the error.  Thread Id information is only present when using the
multi-threaded version of XSB, and even in that version is not present
in all error terms.  

\ourmoditem{xsb\_error\_get\_backtrace(+Error,-Backtrace)}
{xsb\_error\_get\_backtrace/2}{error\_handler} 
%
Obtains the backtrace --- the stack of the forward continuations in
the execution stack at the time the error was thrown.  Backtraces are
present by default in all XSB system error terms, and are described in
Section~\ref{sec:backtrace}.

%\ourmoditem{close\_open\_tables}{close\_open\_tables/0}{machine}
%
%Removes table data structures for all incomplete tables, but does not
%affect any complete tables.  In \version{} this predicate should only
%be used to handle exceptions in {\tt default\_user\_error\_handler/1}.
%In addition, for the multi-threaded engine, this predicate unlocks any
%system mutexes held by the thread calling this predicate.

\end{description}

%----------------------------------------------------------------------------
\section{Convenience Predicates}

The following convenience predicates are provided to make a commonly
used check and to throw an ISO error if the check is not satisfied;
some are written directly in C for speed.  All these predicates must
be imported from the module {\tt error\_handler}, which also contains
provides a few other specialized checks.

\begin{description}
\ourmoditem{check\_acyclic(?Term,+Predicate,+Arg)}{check\_acyclic/3}{error\_handler}
%
Checks that {\tt Term} is acyclic.  If so, the predicate succeeds;
if not it throws a miscellaneous error.

\ourmoditem{check\_atom(?Term,+Predicate,+Arg)}{check\_atom/3}{error\_handler}
%
Checks that {\tt Term} is an atom.  If so, the predicate succeeds;
if not it throws a type error.

\ourmoditem{check\_callable(?Term,+Predicate,+Arg)}{check\_callable/3}{error\_handler}
%
Checks that {\tt Term} is callable.  If so, the predicate succeeds; if
not it throws a type error.

\ourmoditem{check\_ground(?Term,+Predicate,+Arg)}{check\_ground/3}{error\_handler}
%
Checks that {\tt Term} is ground.  If so, the predicate succeeds;
if not it throws an instantiation error.

\ourmoditem{check\_integer(?Term,+Predicate,+Arg)}{check\_integern/3}{error\_handler}
%
Checks that {\tt Term} is an integer.  If so, the predicate succeeds;
if not it throws a type error.

\ourmoditem{check\_nonvar(?Term,+Predicate,+Arg)}{check\_nonvar/3}{error\_handler}
%
Checks that {\tt Term} is not a variable.  If not, the predicate succeeds;
if {\tt Term} is a variable,  it throws an instantiation error.

\ourmoditem{check\_nonvar\_list(?Term,+Predicate,+Arg)}{check\_nonvar\_list/3}{error\_handler}
%
Checks that {\tt Term} is a list, each of whose elements is ground.
If so, the predicate succeeds; if not it throws an instantiation
error.
	    
%\ourmoditem{check\_one\_thread(+Operation,+Object\_Type,+Predicate)}{check\_on%e\_thread/3}{error\_handler}
%
%In the multi-threaded engine, {\tt check\_one\_thread/3} checks that
%there is only one active thread: if not, a miscellaneous error is
%thrown indicating that {\tt Operation} is not permitted on {\tt
%  ObjectType} as called by {\tt Predicate}, when more than one thread
%is active.  This check provides a convenient way to allow inclusion of
%certain operations that are difficult to make thread-safe by other
%means.

%In the single-threaded engine this predicate always succeeds.

\ourmoditem{check\_stream(?Stream,+Predicate,+Arg)}{check\_stream/3}{error\_handler}
%
Checks that {\tt Stream} is a stream.  If so, the predicate succeeds;
if not it throws an instantiation error~\footnote{The representation
of streams in XSB is subject to change.}.

\ourmoditem{check\_var(?Term,+Predicate,+Arg)}{check\_var/3}{error\_handler}
%
Checks that {\tt Term} is a variable.  If so, the predicate succeeds;
if not it throws an instantiation error.

\end{description}

%----------------------------------------------------------------------------
\section{Backtraces}
\label{sec:backtrace}
\index{permanent variables}

\index{Prolog flags!{\tt backtrace\_on\_error}} 
%
Displaying a backtrace of the calling context of an error in addition
to an error message can greatly expedite debugging.  For XSB's default
error handler, backtraces are printed out by default, a behavior that
can be overridden for a given thread by the command: {\tt
  set\_prolog\_flag(backtrace\_on\_error,off)}.  For users who write
their own error handlers, the following predicates can be used to
manipulate backtraces.

It is important to note that Prolog backtraces differ in a significant
manner from backtraces obtained from other languages, such as C
backtraces produced by GDB.  This is because a Prolog backtrace
obtains forward continuations from the local environment stack, and in
the WAM, local stack frames are only created when a given clause
requires permanent variables -- otherwise these stack frames are
optimized away.  The precise conditions for optimizing away a local
stack frame require an understanding of the WAM (and of a specific
compiler).  However in general, longer clauses with many variables
require a local stack frame and their forward continuations will be
displayed, while shorter clauses with fewer variables do not and their
forward continuations will not be displayed.

\begin{description}
\ourmoditem{xsb\_backtrace(-Backtrace)}{xsb\_backtrace/1}{machine}
%
Upon success {\tt Backtrace} is bound to a structure indicating the
forward continuations for a point of execution.  This structure should
be treated as opaque, and manipulated by one of the predicates below.
%
\ourmoditem{get\_backtrace\_list(+Backtrace,-PredicateList)}{get\_backtrace\_list/2}{error\_handler}
%
Given a backtrace structure, this predicate produces a list of
predicate identifiers or the form {\tt Module:Predicate/Arity}.  This
list can be manipulated as desired by error handling routines.
%
\ourmoditem{print\_backtrace(+Backtrace)}{print\_backtrace/1}{error\_handler}
%
 This predicate, which is used by XSB's default error handler, prints
 a backtrace structure to XSB's standard error stream.
\end{description}
\index{exceptions|)}

When XSB generates a memory exception {\em at the OS level} (e.g., a
segmentation violation or bus error) it prints out a backtrace and
exits.  This should be caused only by a bug in XSB or included C code.
The first predicate in the backtrace that is printed in these
circumstances may be incorrect or redundant.  This is because the
memory structures used to generate the backtrace are not always
completely consistent, and so an interrupt at an unexpected point may
result in the use of somewhat inconsistent information.

