\chapter{Hooks} \label{hooks}

Sometimes it is useful to let the user application catch certain
events that occur during XSB execution. For instance, when the user
asserts or retracts a clause, etc.  XSB has a general mechanism by
which the user program can register \emph{hooks} to handle certain
supported events. All the predicates described below must be imported
from {\tt xsb\_hook}.


\section{Adding and Removing Hooks}

A hook in XSB can be either a 0-ary predicate or a unary predicate.
A 0-ary hook is called without parameters and unary hooks are called with
one parameter. The nature of the parameter depends on the type of the hook,
as described in the next subsection.


\begin{description}
\ourmoditem{add\_xsb\_hook(+HookSpec)}{add\_xsb\_hook/1}{xsb\_hook}

This predicate registers a hook; it must be imported from {\tt xsb\_hook}.
{\tt HookSpec} has the following format:
%%
\begin{quote}
 {\tt
   hook-type(your-hook-predicate(\_))
   }
\end{quote}
%%
or, if it is a 0-ary hook:
%%
\begin{quote}
  {\tt
   hook-type(your-hook-predicate)
   }  
\end{quote}
%%
For instance, 
%%
\begin{verbatim}
    :- add_xsb_hook(xsb_assert_hook(foobar(_))).
\end{verbatim}
%%
registers the hook {\tt foobar/1} as a hook to be called when XSB
asserts a clause. Your program must include
clauses that define {\tt foobar/1}, or else an error will result.

The predicate that defines the hook type must be imported from {\tt
  xsb\_hook}:
%%
\begin{verbatim}
    :- import xsb_assert_hook/1 from xsb_hook.  
\end{verbatim}
%%
or {\tt add\_xsb\_hook/1} will issue an error.

\ourmoditem{remove\_xsb\_hook(+HookSpec)}{remove\_xsb\_hook/1}{xsb\_hook}

Unregisters the specified XSB hook; imported from {\tt xsb\_hook}. For
instance,
%%
\begin{verbatim}
    :- remove_xsb_hook(xsb_assert_hook(foobar(_))).
\end{verbatim}
%%
As before, the predicate that defines the hook type must be imported from
{\tt xsb\_hook}.
\end{description}


\section{Hooks Supported by XSB}

The following predicates define the hook types supported by XSB. They must
be imported from {\tt xsb\_hook}.

\begin{description}
\ourmoditem{xsb\_exit\_hook(\_)}{xsb\_exit\_hook/1}{xsb\_hook} 

These hooks are called just before XSB exits. You can register as many
hooks as you want and all of them will be called on exit (but the order of
the calls is not guaranteed). Exit hooks are all 0-ary and must be registered
as such:
%%
\begin{verbatim}
    :- add_xsb_hook(xsb_exit_hook(my_own_exit_hook)).
\end{verbatim}
%%


\ourmoditem{xsb\_assert\_hook(\_)}{xsb\_assert\_hook/1}{xsb\_hook} 
%
These hooks are called whenever the program asserts a clause. An assert
hook must be a unary predicate, which expects the clause
being asserted as a parameter. For instance,
%%
\begin{verbatim}
    :- add_xsb_hook(xsb_assert_hook(my_assert_hook(_))).
\end{verbatim}
%%
registers {\tt my\_assert\_hook/1} as an assert hook. One can register
several assert hooks and all of them will be called (but the order is
not guaranteed).  After the assert\_hooks are called, the original
assert will be called to add the clause to the database.

\ourmoditem{xsb\_intercept\_assert\_hook(\_)}{xsb\_intercept\_assert\_hook/1}{xsb\_hook} 
%
These hooks are called whenever the program asserts a clause.  (They
are called after all xsb\_assert\_hooks, if any, are called.)  An
intercept\_assert hook must be a unary predicate, which expects a
length 2 list as a parameter.  The first element of the list is the
clause being asserted; the second element is 1 for assert (and
assertz), or 0 for asserta.  For instance,
%%
\begin{verbatim}
:- add_xsb_hook(xsb_intercept_assert_hook(my_intercept_assert_hook(_))).
\end{verbatim}
%%
registers {\tt my\_intercept\_assert\_hook/1} as an intercept\_assert
hook. One can register several intercept\_assert hooks and the first
one that succeeds will apply (the order is not guaranteed).  If any
call to an intercept\_assert\_hook succeeds, then the assert is {\em not}
invoked, so it is expected that the intercepting hook will assert any
desired clause.  If no call to the hook predicate succeeds, then the
assert is invoked.  This allows a user, for example, to re-direct all
asserts to one predicate to another different predicate.

\ourmoditem{xsb\_retract\_hook(\_)}{xsb\_retract\_hook/1}{xsb\_hook} 
%
These hooks are called whenever the program retracts a clause, using
{\tt retract/1} or {\tt retractall/1}. A retract hook must be a unary
predicate, which expects as a parameter a list of the form {\tt
  [Head,Body]}, which represent the head and the body parts of the
clause being retracted. As with assert hooks, any number of retract
hooks can be registered and all of them will be called in some order.
After the retract\_hook has been called, the original retract (or
retractall) is called to perform the retraction.

\ourmoditem{xsb\_intercept\_retractall\_hook(\_)}{xsb\_intercept\_retractall\_hook/1}{xsb\_hook} 
%
These hooks are called whenever the program calls {\tt retractall/1}
on a term.  (They are called after all xsb\_retract\_hooks, if any,
are called.)  An intercept\_retractall hook must be a unary predicate,
which expects the term passed to {\tt retractall/1} as a
parameter. For instance,
%%
\begin{verbatim}
:- add_xsb_hook(xsb_intercept_retractall_hook(my_intercept_retractall_hook(_))).
\end{verbatim}
%%
registers {\tt my\_intercept\_retractall\_hook/1} as an intercept\_retractall
hook. One can register several intercept\_retractall hooks and the first
one that succeeds will apply (the order is not guaranteed).  If any
call to an intercept\_retractall\_hook succeeds, then the retractall is {\em not}
invoked, so it is expected that the intercepting hook will retractall any
desired clause(s).

\end{description}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual1"
%%% End: 
