
\chapter{pcre: Pattern Matching and Substitution Using PCRE}
\label{chap-pcre}

\begin{center}
{\Large {\bf By Mandar Pathak}}
\end{center}

\section{Introduction}

This package employs the PCRE library to enable XSB perform pattern
matching and string substitution based on Perl regular expressions.

\section{Pattern matching}

The {\tt pcre} package provides two ways of doing pattern matching:
first-match mode and bulk-match mode. The syntax of the {\tt pcre:match/4}
predicate is:

\begin{verbatim}
?- pcre:match(+Pattern, +Subject, -MatchList, +Mode).
\end{verbatim}

To find only the first match, the {\tt Mode}  parameter must be set to the atom
{\tt one}. To find all matches, the {\tt Mode}  parameter is set to the atom
{\tt bulk}. The result of the matching is returned as a list of 
terms of the form

\begin{center}
\texttt{match}(\textit{Match},\textit{Prematch},\textit{Postmatch},[\textit{Subpattern1, Subpattern2,\ldots}])
\end{center}

The {\tt Pattern} and the {\tt Subject} arguments of {\tt pcre:match} must
be XSB atoms. If there is a match in the subject, then the result is
returned as a list of the \texttt{match(...)}-elements
shown above. \textit{Match} refers to the
substring that matched the entire pattern. \textit{Prematch} contains part
of the subject-string that precedes the matched substring.
\textit{Postmatch} contains part of the subject following the matched
substring. The list of subpatterns (the 4-th argument of the {\tt match}
data structure) corresponds to the substrings that matched the
parenthesized expressions in the given pattern. For example:

\begin{verbatim}
?- pcre:match('(\d{5}-\d{4})\ [A-Z]{2}',
	'Hello12345-6789 NYwalk', X, one).
X = [match(12345-6789 NY,Hello,walk,[12345-6789])]
\end{verbatim}

In this example, the mode argument is \texttt{one} so only one match is
returned, 
the match found for the substring `12345-6789 NY'. The
prematch is `Hello' and the postmatch is `walk'. The substring
\textit{`12345-6789'} matched the parenthesized expression $( \backslash d
\lbrace 5 \rbrace - \backslash d \lbrace 4 \rbrace )$ and hence it is
returned as part of the subpatterns list.

Consider another example, one where all matches are returned:
%%
\begin{verbatim}
?- pcre:match('[a-z]+@[a-z]+\.(com|net|edu)', 
              'a@b.com@c.net@d.edu', X, bulk).
X = [match(a@b.com,,@c.net@d.edu,[com]),
     match(com@c.net,a@b.,@d.edu,[net]),
     match(om@c.net,a@b.c,@d.edu,[net]),
     match(m@c.net,a@b.co,@d.edu,[net]),
     match(net@d.edu,a@b.com@c.,,[edu]),
     match(et@d.edu,a@b.com@c.n,,[edu]),
     match(t@d.edu,a@b.com@c.ne,,[edu])]
\end{verbatim}

This example uses the bulk match mode of the {\tt pcre\_match/4}
predicate to find
all possible matches that resemble a very basic email address. In case
there is no prematch or postmatch to a matched substring, an empty string
is returned.

In general, there can be any number of parenthesized sub-patterns in a given
pattern and the subpattern match-list in the 4-th argument of the {\tt
  match} data structure can have 0, 1, 2, or more elements. 


\section{String Substitution}


The {\tt pcre} package also provides a way to perform string substitution
via the {\tt pcre:substitute/4}  predicate. It has the following syntax:

\begin{verbatim}
?- pcre:substitute(+Pattern, +Subject, +Substitution, -Result).
\end{verbatim}

\textit{Pattern} is the regular expression against which \textit{Subject}
is matched. Each match found is then replaced by the \textit{Substitution},
and the result is returned in the variable \textit{Result}. Here,
\textit{Pattern}, \textit{Subject} and \textit{Substitution} have to be XSB
atoms whereas \textit{Result} must be an unbound variable. The following
example illustrates the use of this predicate:

\begin{verbatim}
?- pcre:substitute(is,'This is a Mississippi issue', was, X).
X = Thwas was a Mwasswassippi wassue
\end{verbatim}

Note that the predicate {\tt pcre:substitute/4} always works in the bulk
mode. If one needs to substitute only \emph{one} occurrence of a pattern,
this is easy to do using the {\tt pcre:match/4} predicate. For instance, if
one wants to replace the third occurrence of ``{\tt is}'' in the above
string, we could issue the query
%% 
\begin{verbatim}
?- pcre:match(is,'This is a Mississippi issue',X,bulk).
\end{verbatim}
%% 
take the third element in the returned list, i.e., 
%% 
\begin{verbatim}
    match(is,'This is a M','sissippi issue',[])
\end{verbatim}
%% 
and then concatenate the \emph{Prematch} in the above \texttt{match(...)}
(i.e., \texttt{'This is a M'}) with the substitute string
(i.e., {\tt 'was'}) and the \emph{Postmatch} (i.e., \texttt{'sissippi issue'}). 

Additional examples of the use of the {\tt pcre} package can be found in
the XSB distribution, in the file \texttt{\$XSBDIR/examples/pcretest.P}.


\section{Installation and configuration}

XSB's {\tt pcre} package requires that the {\tt PCRE} library is installed.
For Windows, the {\tt PCRE} library files are included with the
XSB installation. For Linux and Mac, the PCRE and the PCRE-development
packages must be installed using the distribution's package manager.
The names of these packages might differ from one Linux distribution to
the next. For instance, in Ubuntu, these libraries might be called
\emph{libpcre3} and \texttt{libpcre3-dev}. In contrast, Fedora uses the
names \texttt{pcre} and \texttt{pcre-devel}.  
On the Mac, these packages live in the Homebrew add-on, which
must be installed separately.

\subsection{Configuring for Linux, Mac, and other Unices}

In the unlikely case that your Linux distribution
does not include PCRE as a package they
must be downloaded and built manually. Please visit 
%%
\begin{quote}
http://www.pcre.org/ 
\end{quote}
%%
to download the latest distribution and follow the instructions given with
the package.

To configure {\tt pcre} on Linux, Mac, or on some other Unix variant,
switch to the {\tt XSB/build} directory and type:

\begin{verbatim}
    cd ../packages/pcre
    ./configure
    ./makexsb
\end{verbatim}


\subsection{Configuring for Windows}

If your installation of XSB is not configured with PCRE,
you will need Microsoft \texttt{nmake} installed. 
Change to the top XSB directory and type:
%%
\begin{verbatim}
    cd packages\pcre\cc
    nmake /f NMakefile.mak      <-- if you have the 32 bit version of XSB
    nmake /f NMakefile64.mak    <-- if you have the 64 bit version of XSB
\end{verbatim}
%%
This builds the DLL required by XSB's {\tt pcre} package on Windows. To
make sure that the build went ahead smoothly, open the directory
%%
\begin{verbatim}
{XSB_DIR}\config\x86-pc-windows\bin   <-- if using the 32 bit XSB
{XSB_DIR}\config\x64-pc-windows\bin   <-- if using the 64 bit XSB
\end{verbatim}
%%
and verify that the file {\tt pcre4pl.dll} exists there.

Once the package has been configured, it must be loaded before it can be used:
%%
\begin{verbatim}
?- [pcre].
\end{verbatim}
%%



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual2"
%%% End: 
