\chapter{Syntax} \label{Syntax}
%==============================

The syntax of XSB is based on ISO Prolog~\cite{ISO-Prolog}, although
it lacks a few of the ISO standard's somewhat arcane features.
Beginning with \version, XSB supports Unicode through UTF-8 atoms as
described in Section~\ref{sec:atoms}.  XSB's reader also contains extensions to
support HiLog~\cite{ChKW93}, which adds certain features of
second-order syntax to Prolog.

\section{Terms} \label{TermSyntax}
%=================================
The data objects of the HiLog language are called {\em terms}.
A {\em HiLog term} can be constructed from any logical symbol or a term
followed by any finite number of arguments.  In any case, a {\em term}
is either a {\em constant}, a {\em variable}, or a {\em compound term}.

A {\em constant} is either a {\em number} (integer or floating-point)
or an {\em atom}~\footnote{This Prolog usage contradicts the usage of
  the word ``atom'' in logic as short for ``atomic formula''.}  Constants
are definite elementary objects, and correspond to proper nouns in
natural language.

\subsection{Integers}
\index{syntax!integers}
\index{syntax!integers!binary}
\index{syntax!integers!octal}
\index{syntax!integers!hexidecimal}
\subsubsection{ISO Integers}
%
The printed form of an integer normally consists of a sequence of
digits optionally preceded by a minus sign ({\tt '-'}), interpreted,
of course, as base $10$ integers.  It is also possible to enter
integers in other bases: 
\begin{itemize}
\item {\tt 0b}{\em nnn} represents an integer in base 2, e.g., 
\begin{verbatim}
                        | ?- X = 0b110.

                        X = 6
\end{verbatim}

\item {\tt 0o}{\em nnn} represents an integer in base 8, e.g., 
\begin{verbatim}
                        | ?- X = 0o110.

                        X = 72
\end{verbatim}

\item {\tt 0x}{\em nnn} represents an integer in base 16, e.g., 
\begin{verbatim}
                        | ?- X = 0x110.

                        X = 272
\end{verbatim}
\end{itemize}
\index{character code constants}
Character code constants are integers of the form {\tt 0'}{\em nnn},
where nnn is the decimal form of any UTF-8 codepoint.  E.g.,
\begin{verbatim}
                        | ?- 0'A = X

                        X = 65
\end{verbatim}
Escape characters (cf. Section~\ref{sec:atoms}) can be written
similarly (if this is ever needed):
\begin{verbatim}
                        | ?- 0'\n = X

                        X = 10
\end{verbatim}

\subsubsection{Other Integer Representations}

It is also possible to enter integers in bases $2$ through $36$; this
can be done by preceding the digit sequence by the base (in decimal)
followed by an apostrophe ({\tt '}).  If a base greater than $10$ is
used, the characters {\tt A-Z} or {\tt a-z} are used to stand for
digits greater than $9$.

Using these rules, examples of valid integer representations in XSB are:
\begin{verbatim}
           1    -3456    95359    9'888    16'1FA4    -12'A0    20'
\end{verbatim}
representing respectively the following integers in decimal base:
\begin{verbatim}
           1    -3456    95359     728       8100      -120      0
\end{verbatim}

Note that the following:
\begin{verbatim}
                   +525     12'2CF4     37'12     20'-23
\end{verbatim}
are not valid integers of XSB.

Character code constants, mentioned above, can be seen as integers in ``base zero''.

%\begin{CJK}{UTF8}
%今
%\end{CJK}


\index{syntax!floats}
\subsection{Floating-point Numbers}
XSB supports ISO floating-point numbers, which consist of a sequence
of digits with an embedded decimal point, optionally preceded by a
minus sign ({\tt '-'}), and optionally followed by an exponent
consisting of uppercase or lowercase {\tt 'E'} and an optionally
signed base $10$ integer.

Using these rules, examples of floating point numbers are:
\begin{verbatim}
              1.0    -34.56    817.3E12    -0.0314e26    2.0E-1
\end{verbatim}
Note that in any case there must be at least one digit before, and one digit
after, the decimal point.


\index{syntax!atoms}
\index{syntax!escaped characters}
\subsection{Atoms} \label{sec:atoms}
\index{Unicode!UTF-8}
An atom consists of a sequence of characters that follow the following rules.

\begin{itemize}
\item {\em Non-quoted Atoms} begin with the ASCII character {\tt
  a}-{\tt z} and are followed by a sequence of {\em ISO alphanumeric characters}: {\tt
  a}-{\tt z}, {\tt A}-{\tt Z}, {\tt 0}-{\tt 9}, and underscore {\tt \_}.
%
\item {\em Quoted Atoms} begin and end with the ASCII character \verb|'| and may contain any sequence of 
\begin{itemize}
\item Printable UTF-8 characters

\item Meta-escaped quotes.  E.g., 
\begin{verbatim}
                        | ?- X = 'a''b'.

                        X = a'b
\end{verbatim}
(Unfortunately, the current version of XSB does not support escaped
quotes (\verb|\'|).)
\item ISO escape characters and sequences
\begin{itemize}
\item \verb|\b| the newline character (ASCII 7).

\item \verb|\b| the newline character (ASCII 8).

\item \verb|\f| the form feed character (ASCII 12).

\item \verb|\n| the newline character (ASCII 10).

\item \verb|\r| the carriage return character (ASCII 13).

\item \verb|\t| a tab character (ASCII 9).

\item \verb|\v| a vertical tab character (ASCII 11).

\item Octal escapes of the form \verb|\nnn\|, where {\tt nnn} is the octal number corresponding to an ASCII code.  E.g., 
\begin{verbatim}
                        | ?- write('\60\').
                        0
\end{verbatim}
\item Hexidecimal escapes of the form \verb|\xnn|, where {\tt nn} is
  the hexidecimal number corresponding to an ASCII code~\footnote{The
    current version of XSB differs from the ISO specification in that
    hexidecimal escapes do not have a trailing slash.}.  .  E.g.,
\begin{verbatim}
                        | ?- write('\30\').
                        0
\end{verbatim}
\end{itemize}
\item UTF-8 escape sequences have the form \verb|\unnnn| where {\em
  nnnn} is the hexidecimal number corresponding to a UTF-8 codepoint.  
\end{itemize}
\item {\em Operator-based Atoms} are defined as any sequence from the following set
  of characters (except of the sequence {\tt '/*'}, which begins a
  comment):
      \begin{verbatim}
                     + - * / \ ^ < > = ` ~ : . ? @ # &
      \end{verbatim}
Examples of such atoms are:
\begin{verbatim}
               ^=..   ::=   ===
\end{verbatim}

\item {\em Special Atoms} are
      \begin{verbatim}
                                 !  ;  []  {}
      \end{verbatim}
      Note that the bracket pairs are special. While {\tt '[]'} and
      {\tt '$\{\}$'} are atoms, {\tt '['}, {\tt ']'}, {\tt '$\{$'},
      and {\tt '$\}$'} are not~\footnote{The form {\tt [X]} is a
        special notation for lists (see Section~\ref{Lists}), while
        the form {\tt $\{$X$\}$} is just ``syntactic sugar'' for the
        term {\tt '$\{\}$'(X)}.}.
\end{itemize}



\subsection{Variables}
Variables may be written as any sequence of (ASCII) ISO alphanumeric
characters  beginning with either a capital
letter or {\tt '\_'}.  For example:
\begin{verbatim}
                      X   HiLog   Var1   _3   _List
\end{verbatim}

If a variable is referred to only once in a clause, it does not need
to be named and may be written as an {\em anonymous variable},
represented by a single underscore character {\tt '\_'}.  Any number
of anonymous variables may appear in a clause; all of these variables
are read as distinct variables.  


\subsection{Compound Terms}
Like in Prolog, the structured data objects of HiLog are {\em compound terms}
(or {\em structures}).  The external representation of a HiLog compound term
comprises a {\em functor} (called the {\em principal functor} or the
{\em name} of the compound term) and a sequence of one or more terms called
{\em arguments}.  Unlike Prolog where the functor of a term must be an atom,
in HiLog the functor of a compound term {\em can be any valid HiLog term}.
This includes numbers, atoms, variables or even compound terms.  Thus, since
in HiLog a compound term is just a term followed by any finite number of
arguments, all the following are valid external representations of HiLog
compound terms: 
\label{some_compound_terms}
\begin{verbatim}
          foo(bar)             prolog(a, X)              hilog(X)       
       123(john, 500)        X(kostis, sofia)         X(Y, Z, Y(W))   
      f(a, (b(c))(d))       map(double)([], [])     h(map(P)(A, B))(C)
\end{verbatim}

Like a functor in Prolog, a functor in HiLog can be characterized by
its {\em name} and its {\em arity} which is the number of arguments this
functor is applied to.  For example, the compound term whose principal functor
is {\tt 'map(P)'} of arity 2, and which has arguments {\tt L1}, and {\tt L2},
is written as:
\begin{verbatim}
                            map(P)(L1, L2)
\end{verbatim}

As in Prolog, when we need to refer explicitly to a functor we will normally
denote it by the form $Name/Arity$.  Thus, in the previous example, the functor
{\tt 'map(P)'} of arity 2 is denoted by:
\begin{verbatim}
                                 map(P)/2
\end{verbatim}
Note that a functor of arity 0 is represented as an atom.

In Prolog, a compound term of the form $p(t_1, t_2, \ldots, t_k)$ is usually
pictured as a tree in which every node contains the name $p$ of the functor
of the term and has exactly $k$ children each one of which is the root of the
tree of terms $t_1, t_2, \ldots, t_k$.

For example, the compound term
\begin{verbatim}
                 s(np(kostis), vp(v(loves), np(sofia)))
\end{verbatim}
would be pictured as the following tree:

\begin{minipage}{4.0in}
\begin{verbatim}
                                  s
                                /   \
                             np       vp
                             |       /  \
                             |      v     np
                             |      |     |
                          kostis  loves  sofia
\end{verbatim}
\end{minipage}

\noindent
The principal functor of this term is {\tt s/2}.  Its two arguments are also
compound terms.  In illustration, the principal functor of the second
argument is {\tt vp/2}.

Likewise, any external representation of a HiLog compound term
$t(t_1, t_2, \ldots, t_k)$ can be pictured as a tree in which every node
contains the tree representation of the name $t$ of the functor of the term
and has exactly $k$ children each one of which is the root of the tree of
terms $t_1, t_2, \ldots, t_k$.

Sometimes it is convenient to write certain functors as {\em operators}.
{\em Binary functors} (that is, functors that are applied to two arguments)
may be declared as {\em infix operators}, and {\em unary functors} (that is,
functors that are applied to one argument) may be declared as either 
{\em prefix or postfix operators}.  
Thus, it is possible to write the following:
\begin{verbatim}
                    X+Y     (P;Q)     X<Y      +X     P;
\end{verbatim}
More about operators in HiLog can be found in section~\ref{Operators}.


\subsection{Lists}\label{Lists}
As in Prolog, lists form an important class of data structures in HiLog.
They are essentially the same as the lists of Lisp: a list is either the atom
{\tt '[]'}, representing the empty list, or else a compound term with functor
{\tt '.'}  and two arguments which are the head and tail of the list
respectively, where the tail of a list is also a list.
Thus a list of the first three natural numbers is the structure:
\begin{verbatim}
                                  .
                                 / \
                                1    .
                                    / \
                                   2    .
                                       / \
                                      3   []
\end{verbatim}
which could be written using the standard syntax, as:
\begin{verbatim}
                             .(1,.(2,.(3,[])))
\end{verbatim}
but which is normally written in a special list notation, as:
\begin{verbatim}
                                  [1,2,3]
\end{verbatim}
Two examples of this list notation, as used when the tail of a list is a
variable, are:
\begin{verbatim}
                       [Head|Tail]      [foo,bar|Tail]
\end{verbatim}
which represent the structures:
\begin{verbatim}
                            .                .
                           / \              / \
                       Head   Tail        foo   .
                                               / \
                                             bar  Tail
\end{verbatim}
respectively.

Note that the usual list notation {\tt [H|T]} does not add any new power
to the language; it is simply a notational convenience and improves
readability. The above examples could have been written equally well as:
\begin{verbatim}
                      .(Head,Tail)      .(foo,.(bar,Tail))
\end{verbatim}

For convenience, a further notational variant is allowed for lists of
integers that correspond to UTF-8< character codes.  Lists written in this
notation are called {\em strings}.  For example,
\begin{verbatim}
                            "I am a HiLog string"
\end{verbatim}
represents exactly the same list as:
\begin{verbatim}
    [73,32,97,109,32,97,32,72,105,76,111,103,32,115,116,114,105,110,103]
\end{verbatim}


\section{From HiLog to Prolog} \label{HiLog2Prolog}
%==================================================
From the discussion about the syntax of HiLog terms, it is clear that the
HiLog syntax allows the incorporation of some higher-order constructs in
a declarative way within logic programs.  As we will show in this section,
HiLog does so  while retaining a clean first-order declarative semantics.
The semantics of HiLog is first-order, because every HiLog term (and formula)
is automatically {\em encoded (converted)} in predicate calculus in the way
explained below.

Before we briefly explain the encoding of HiLog terms, let us note that the 
HiLog syntax is a simple (but notationally very convenient) encoding for Prolog
terms, of some special form.  In the same way that in Prolog:
\begin{center}
{\tt	1 + 2}
\end{center}
is just an (external) shorthand for the term:
\begin{center}
{\tt  +(1, 2)} 
\end{center}
in the presence of an infix operator declaration for {\tt +} 
(see section~\ref{Operators}), so:
\begin{center}
{\tt  X(a, b)}
\end{center}
is just an (external) shorthand for the Prolog compound term:
\begin{center}
{\tt    apply(X, a, b)}
\end{center}
Also, in the presence of a {\tt hilog} declaration (see
section~\ref{other-directives}) for {\tt h}, the HiLog term whose external
representation is:
\begin{center}
{\tt  h(a, h, b)} 
\end{center}
is a notational shorthand for the term:
\begin{center}
{\tt apply(h, a, h, b)}
\end{center}
Notice that even though the two occurrences of {\tt h} refer to the same 
symbol, only the one where {\tt h} appears in a functor position is encoded
with the special functor {\tt apply/}$n, n \geq 1$.

The encoding of HiLog terms is performed based upon the existing declarations
of {\em hilog symbols}.  These declarations (see section~\ref{other-directives}),
determine whether an atom that appears in a functor position of an external 
representation of a HiLog term, denotes a functor or the first argument of a 
set of special functors {\tt apply}.  The actual encoding is as follows:
\begin{itemize}
\item	The encoding of any variable or parameter symbol (atom or number) that
	does not appear in a functor position is the variable or the symbol
	itself.
\item	The encoding of any compound term {\tt t} where the functor {\em f}
	is an atom that is not one of the hilog symbols (as a result of a
	previous {\tt hilog} declaration), is the compound term that has
	{\em f} as functor and has as arguments the encoding of the arguments
	of term {\em t}.  Note that the arity of the compound term that results
	from the encoding of {\em t} is the same as that of {\em t}.
\item	The encoding of any compound term {\tt t} where the functor {\em f}
	is either not an atom, or is an atom that is a hilog symbol, is a 
	compound term that has {\tt apply} as functor, has first argument
	the encoding of {\em f} and the rest of its arguments are obtained
	by encoding of the arguments of term{\em t}.  Note that in this case
	the arity of the compound term that results from the encoding of
	{\em t} is one more than the arity of {\em t}.
\end{itemize}

Note that the encoding of HiLog terms described above, implies that even
though the HiLog terms:
\begin{center}
\begin{minipage}{1.0in}
\begin{verbatim}
	p(a, b)
	h(a, b)
\end{verbatim}
\end{minipage}
\end{center}
externally appear to have the same form, in the presence of a {\tt hilog}
declaration for {\tt h} but not for {\tt p}, they are completely different.
This is because these terms are shorthands for the terms whose internal 
representation is: 
\begin{center}
\begin{minipage}{1.2in}
\begin{verbatim}
	    p(a, b)
	apply(h, a, b)
\end{verbatim}
\end{minipage}
\end{center}
respectively.  Furthermore, only {\tt h(a,b)} is unifiable with the HiLog term
whose external representation is {\tt X(a, b)}.

We end this short discussion on the encoding of HiLog terms with a small
example that illustrates the way the encoding described above is being done.
Assuming that the following declarations of parameter symbols have taken place,
\begin{center}
\begin{minipage}{1.6in}
\begin{verbatim}
 :- hilog h.
 :- hilog (hilog).
\end{verbatim}
\end{minipage}
\end{center}
before the compound terms of page~\pageref{some_compound_terms} were
read by XSB, the encoding of these terms in predicate calculus using
the described transformation is as follows:
\begin{center}
\begin{minipage}{4.5in}
\begin{verbatim}
      foo(bar)                    prolog(a,X) 
   apply(hilog,X)             apply(123,john,500)
apply(X,kostis,sofia)       apply(X,Y,Z,apply(Y,W))
  f(a,apply(b(c),d))       apply(map(double),[],[])  
        apply(apply(h,apply(map(P),A,B)),C)
\end{verbatim}
\end{minipage}
\end{center}


\section{Operators} \label{Operators}
%====================================
From a theoretical point of view, operators in Prolog are simply a notational
convenience and add absolutely nothing to the power of the language.
For example, in most Prologs {\tt '+'} is an infix operator, so
\begin{verbatim}
                                 2 + 1
\end{verbatim}
is an alternative way of writing the term {\tt +(2, 1)}.  That is, {\tt 2 + 1}
represents the data structure:
\begin{verbatim}
                                   +
                                  / \
                                 2   1
\end{verbatim}
and not the number 3.  (The addition would only be performed if the structure
were passed as an argument to an appropriate procedure, such as {\tt is/2}).
% tls we havent done this yet. described in section~\ref{Arithmetic}).

However, from a practical or a programmer's point of view, the existence of
operators is highly desirable, and clearly handy.

Prolog syntax allows operators of three kinds: {\em infix}, {\em prefix}, and
{\em postfix}.  An {\em infix} operator appears between its two arguments,
while a {\em prefix} operator precedes its single argument and a {\em postfix}
operator follows its single argument.

Each operator has a precedence, which is an integer from 1 to 1200.  The
precedence is used to disambiguate expressions in which the structure of the
term denoted is not made explicit through the use of parentheses.  The
general rule is that the operator with the highest precedence is the
principal functor.  Thus if {\tt '+'} has a higher precedence than {\tt '/'},
then the following
\begin{verbatim}
                           a+b/c     a+(b/c)
\end{verbatim}
are equivalent, and both denote the same term {\tt +(a,/(b,c))}. Note that
in this case, the infix form of the term {\tt /(+(a,b),c)} must be written
with explicit use of parentheses, as in:
\begin{verbatim}
                               (a+b)/c
\end{verbatim}

If there are two operators in the expression having the same highest
precedence, the ambiguity must be resolved from the {\em types} (and 
the implied {\em associativity}) of the operators.  The possible types
for an infix operator are
\begin{verbatim}
                          yfx     xfx     xfy
\end{verbatim}
Operators of type {\tt 'xfx'} are not associative.  Thus, it is required that
both of the arguments of the operator must be subexpressions of lower
precedence than the operator itself; that is, the principal functor of each
subexpression must be of lower precedence, unless the subexpression is written
in parentheses (which automatically gives it zero precedence).

Operators of type {\tt 'xfy'} are {\em right-associative}:  only the
first (left-hand) subexpression must be of lower precedence; the right-hand
subexpression can be of the same precedence as the main operator.
{\em Left-associative} operators (type {\tt 'yfx'}) are the other way around.

An atom named {\tt Name} can be declared as an operator of type {\tt
Type} and precedence {\tt Precedence} by the command;

\begin{description}
\isoitem{op(+Precedence,+Type,+Name)}{op/3}
\end{description}
%
The same command can be used to redefine one of the predefined XSB
operators (obtainable via {\tt current\_op/3}).  However, it is not
allowed to alter the definition of the comma ({\tt ','}) operator.  An
operator declaration can be cancelled by redeclaring the {\tt Name}
with the same {\tt Type}, but {\tt Precedence} 0.

As a notational convenience, the argument {\tt Name} can also be a list of
names of operators of the same type and precedence.

It is possible to have more than one operator of the same name, so
long as they are of different kinds: infix, prefix, or postfix.  An
operator of any kind may be redefined by a new declaration of the same
kind.  For example, the built-in operators {\tt '+'} and {\tt '-'} are
as if they had been declared by the command:
\begin{verbatim}
                       :- op(500, yfx, [+,-]).
\end{verbatim}
so that:
\begin{verbatim}
                                  1-2+3
\end{verbatim}
is valid syntax, and denotes the compound term:
\begin{verbatim}
                                 (1-2)+3
\end{verbatim}
or pictorially:
\begin{verbatim}
                                    +
                                   / \
                                  -   3
                                 / \
                                1   2
\end{verbatim}

In XSB, the list functor {\tt '.'/2} is one of the standard operators,
that can be thought as declared by the command:
\begin{verbatim}
                          :- op(661, xfy, .).
\end{verbatim}
So, in XSB,
\begin{verbatim}
                                  1.2.[]
\end{verbatim}
represents the structure
\begin{verbatim}
                                    .
                                   / \
                                  1   .
                                     / \
                                    2   []
\end{verbatim}
Contrasting this picture with the picture above for {\tt 1-2+3} shows
the difference between {\tt 'yfx'} operators where the tree grows to
the left, and {\tt 'xfy'} operators where it grows to the right.  The
tree cannot grow at all for {\tt 'xfx'} type operators.  It is simply
illegal to combine {\tt 'xfx'} operators having equal precedences in
this way.

If these precedence and associativity rules seem rather complex, remember
that you can always use parentheses when in any doubt.

In \version{} of XSB the possible types for prefix operators are:
\begin{verbatim}
                      fx       fy       hx       hy
\end{verbatim}
and the possible types for postfix operators are:
\begin{verbatim}
                               xf       yf
\end{verbatim}

We end our discussion about operators by mentioning that prefix
operators of type {\tt hx} and {\tt hy} are {\em proper HiLog
  operators}.  The discussion of proper HiLog operators and their
properties is deferred for the manual of a future version.
\footnote{As a known bug, XSB's reader cannot properly read an
  operator defined as both a prefix and an infix operator.  For
  instance the declaration of both {\tt :- op(1200,xf,'$<=$').} and 
{\tt :- op(1200,xfx,'$<=$').} will lead to a syntax error.}

\section{Canonical Syntax} \label{sec:canonical}
\index{canonical format}%
Prolog syntax depends heavily on operators even if a user does not
define them explicitly.  For instance, rules made use of the infix
\mif{} operator together with infix operators for conjunction,
disjunction, the use of parentheses to override priorities and so on.
Within {\tt is/2} arithmetic and logical expressions also rely on
operators.  

This syntactic flexibility is crucial to making Prolog code easy to
read and enjoyable to write.  However, the widesperad use of operators
and their precidance makes file reading slower than it otherwise would
be.  For most files this slowdown is not even noticeable, but for
large knowledge bases of many millions of clauses, reading can add a
perceptible overhead to load time.  This overhead can be avoided when
loading dynamic code if the input file is in {\em canonical format},
i.e., if all of its clauses are in canonical syntax.  Canonical syntax
may also be useful if XSB needs to write results that need to be
parsed by some other language.

Canonical syntax ensures that all terms, atoms, and rules are written
in prefix order.  XSB's version of canonical syntax supports two
exceptions to this: the normal syntax of lists and comma lists are
considered canonical.  For instance, the syntax {\tt [a,b,c]} is
allowed and the fully prefix syntax {\tt '.'(a,'.'(b,c))} is not
requiried.  In addition, canonical format in XSB disallows constructs
that are based on transformations, such as DCGs, HiLog and certain
types of tabling.

Canonical syntax is most easily understood via examples.
The rule:

{\tt 
anc(X,Y,L):- anc(X,Z,L1),par(Z,Y),L is L + 1.
}

has canonical syntax

{\tt 
  :-(anc(X,Y,L),','(anc(X,Z,L1),','(par(Z,Y),is(L,+(L1,1))))).
}

while the rule

{\tt 
  append([X|L1],L2,[X|L3]) :- append(L1,L2,L3))).
}

has canonical syntax

{\tt 
:-(append([X|L1],L1,[X|L3]),append(L1,L2,l3)).
}

where the lists are allowed to use infix notation.

A Prolog file in normal syntax can be converted to canonical format
via the predicate {\tt cvt\_canonical(InFile,OutFile)}.  Prolog rules,
atoms, or terms can be read or written in canonical format by {\tt
  read\_canonical/[1,2]} and {\tt write\_canonical/[1,2]}).
